Evolution of ORM Systems: Formal Foundations

### 2013-01-15 doc/description.tex
added section headers
fixed typos and updated text structure

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented in an object-oriented language, which uses relational database as a data storage is context of this paper. The situation is illustrated in the Figure \ref{fig:appStructure}. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software as a triple:
\begin{align*}
	\mathbf{software} = < Application, Database, \rho > \\
	sw : Applications \times Databases \times ORM \rightarrow Software \\ \\
	\mathbf{application} : Software \rightarrow Application \\
	application(sw(a, d, \rho)) = a \\ \\
	\mathbf{database} : Software \rightarrow Database \\
	database(sw(a, d, \rho)) = d \\
	a \lhd Application, d \lhd Database
\end{align*}
The software definition above introduce the first type used in this paper. The notation $a \lhd Application$ indicates the variable $a$ has the type $Application$.

The software has to be in consistent state so its users could have benefit from its useage. The state of the software is defined to communicate the sate of the software to its user:
\begin{align*}
 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s, \\\\
	\mathbf{state} : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ \\
 	s \lhd Software  
\end{align*}
\subsection{Software Evolution}
\begin{figure}
\centering
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
	\label{fig:evolution}
\end{figure}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software. The evolution consists of change in application and in database as shown in the Figure \ref{fig:evolution}. We focus on situations when the evolution is initiated on the application level (as code refactoring), however the evolution has to be processed on the database level too. It means there is a set of application transformations called $AppOps$ and a set of database transformations called $DbOps$. Because there are needed different parameters for application and database evolution the set of transformation called $E$ which evolves the whole software and a mapping between $E$ and $AppOps$, respectively $DbOps$ is defined. We call this mappings $\Psi$ and $\Phi$:
\begin{align*}
\Psi : E \rightarrow AppOps \\
\Phi : E \rightarrow DbOps
\end{align*}
A subset of $E$ represents an evolution transformation of a software, which mets the first goal of our framework - a single source of data evolution for the whole software.

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
$$\oplus : State \times E \rightarrow State $$
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
\circ : Application \times AppOps \rightarrow Application \\
 \bullet : Database \times DbOps \rightarrow Database
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.

\begin{align*}
 u \oplus t = \begin{cases}
 	\inference{u = Inconsistent \; s}{u} \\ \\
 	\inference{u = Consistent \; s}{state(sw(\Psi(u, t), \Phi(u,t))} 
\end{cases}\\ 
u \lhd State, t \in E, s \lhd Software
\end{align*}
\begin{align*}
 a \circ o = \begin{cases}
 	\inference{!\;consistent(a)}{a} \\ \\
 	\inference{consistent(a)}{a \circ o} 
\end{cases}\\ 
a \lhd Application, o \in AppOps
\end{align*}
\begin{align*}
 d \bullet o = \begin{cases}
 	\inference{!\;consistent(d)}{d} \\ \\
 	\inference{consistent(d)}{d \circ o} 
\end{cases}\\ 
d \lhd Application, o \in DbOps
\end{align*}

\section{Software Model}
The model of an ORM software consists of three parts - an application model, a database model and a ORM. The model of application and database creates a type system for both kinds ORM and evolutionary transformations.

The definitions of all parts assume a set $Label$ exist which contains all possible identifiers and each modeled entity is identified unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model.

\subsection{Application Model}
\label{sec:appModel}
The application model represents a persistent layer of an application created using object oriented language and transformations which can alter the model. 
\subsubsection{Application Structure}
Each instance in an application is defined according to some class - the class defines the structure of its instances and relationships between them an the rest of the application. The instances are stored in the relational database in our case, therefore we define only the structure of classes in our model. 
\begin{align*}
  \mathbf{Application} = < Class* > \\ 
  \mathbf{app} : Class* \rightarrow  Application \\ \\
  \mathbf{classes} : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  \mathbf{class} : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label,a \lhd Applications, cs = Class* 
\end{align*}

\paragraph{Class} Class represents a basic structural unit in the application model. It has a unique name, one or more properties and it can be associated to other classes in the application. 
\begin{align*}
	\mathbf{Class} = <label, Property*, Association*> \\
	\mathbf{cl} : Label \times Property* \times Association* \rightarrow Class \\\\
	\mathbf{name} : Class \rightarrow Label \\
	name(cl(l, props, assocs)) = l \\ \\
	\mathbf{properties} : Class \rightarrow Property* \\
	properties(cl(l, props, assocs)) = props \\ \\
	\mathbf{associations} : Class \rightarrow Association* \\	associations(cl(l, props, assocs)) = assocs \\ \\
\end{align*}
Three more function are defined to obtain information about a class - $primitives$ returns all properties of the class which represent a single value, whereas $collections$ returns all properties representing collections. The last one $associated$ determine if a class is associated by any other class in an application.
\begin{align*}
	\mathbf{primitives} : Class \rightarrow Property* \\	primitives(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
	\mathbf{collections} : Class \rightarrow Property* \\	collections(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
	\mathbf{associated} :  Label \times Application \rightarrow Boolean \\
  	associated(l,a) = \begin{cases}
  		\inference{\exists c \in classes(a) \wedge \exists e \in associations(c) : \\ reference(e) = l
  	}{ true } \\\\
 		false
 	\end{cases} \\ \\
	l \in Label, props = Property*, assocs = Association*,a \lhd Applications
\end{align*}

\paragraph{Property} Property represents a feature of a class which is represented as a primitive type. The property can be mandatory, can have a default value and according to its cardinality it can represent a single value or a collection of values. The $owningClass$ function returns the class in the model which owns the property.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\ \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
	\mathbf{owningClass} : Property \times Application \rightarrow Class  \\ 	
	owningClass(p, a) = \inference{\exists c \in classes(a) \wedge p \in properties(c)}{c}  \\ \\
	l \in Label, t \lhd AppType, val \in Values, n \in \mathbb{N^{*}}, m \in Boolean 
\end{align*}

\paragraph {Association} Assoctation represents a connection between two classes. It has a unique name and the reference is represented by the label of referenced class. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of the association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\ \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\ \\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
	\mathbf{owningClass} : Association \times Application \rightarrow Class  \\
	owningClass(as, a) = \inference{\exists c \in classes(a) \wedge as \in associations(c)}{c}  \\ \\	
	l, ref \in Label,  n_1, n_2 \in \mathbb{N}, as \lhd Association
\end{align*}

@@

\paragraph{Application type} Application type ($AppType$) represents primitive type in the application. There are usually defined types such as String, Integer, Boolean etc. in contrast there is only one type in our model, because we focus on structural and data changes and type casting operations are not important for us. The only App-Type type is called $APPSTRING$.
\begin{align*}
\mathbf{AppType} = APPSTRING
\end{align*}

\subsection{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is:
\begin{align*}
	\mathbf{Database} = < Table*, Row*> \\ \\
	\mathbf{db} : Tables \times Rows \rightarrow Database \\ \\
	\mathbf{tables} : Database \rightarrow Tables \\
	tables(db(tabs, rs)) = tabs \\ \\
	\mathbf{rows} : Database \rightarrow Rows \\
	rows(db(tabs, rs)) = rs \\ \\
	tabs = Table*, rs = Row*, l \in Label,d \lhd Database
\end{align*}
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{align*}
	\mathbf{Table} = <label, primaryKey, Column*, ForeignKey*>\\ \\
	\mathbf{tab} 	: Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
	\mathbf{name} : Table \rightarrow Label  \\
	name(tab(l, pk, cols, fks)) = l \\ \\
	\mathbf{primaryKey} : Table \rightarrow PrimaryKeys  \\
	primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
	\mathbf{columns} : Table \rightarrow Columns  \\
	columns(tab(l, pk, cols, fks)) = cols \\ \\
	\mathbf{foreignKeys} : Table \rightarrow ForeignKeys  \\
	foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk \lhd PrimaryKey, cols = Column*, fks = ForeignKey*
\end{align*}

\paragraph{Column} columns define possible data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\ \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
	\mathbf{owningTable} : Column \times Database \rightarrow Table  \\
	owningTable(c, d) = \inference{\exists tab \in tables(d) \wedge c \in columns(tab)}{tab} \\ \\
l \in Label, t \lhd DbType, val \in Values,\\ cons \lhd Constraints, c \lhd Column,d \lhd Database
\end{align*}

\paragraph{Foreign key} Foreign key is a reference to another table's primary key, it has an unique name and it can be constrained. The value of a foreign key is $\varnothing$ or a non-zero natural number.
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ \\
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
	l, tRef \in Label, cons \lhd Constraint*,d \lhd Database, fk \lhd ForeignKey
\end{align*}

\paragraph{Primary key} Primary key is unambiguous identifier of a record in a table. The primary key is always provided by the database as a non-zero natural number. Primary key is always defined with constraints $NOTNULL$ and $UNIQUE$. 
\begin{align*}
	\mathbf{PrimaryKey} =  < label > 	\\
	\mathbf{pk} : Label \rightarrow PrimaryKey
\end{align*}

\paragraph{Data Types} Database data types $DbType$ represents primitive types in the database. There are usually defined types such as Varchar, Integer, Boolean etc. There is only one type defined in the model $DBSTRING$ as a default type of values in columns.
\begin{align*}
	\mathbf{DbType} = DBSTRING
\end{align*}

\paragraph{Constraints} There are two types of constraints defined in the model. Both constraints are column constraints - first constraint defines unique records in a column, second constraint defines non-empty columns.
\begin{align*}
	\mathbf{Constraint} = NOTNULL \; | \; UNIQUE 
\end{align*}

@@

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete values of concrete columns. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pkValue, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, n, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, n, ps)) = ps \\ \\
	\mathbf{pkValue} : Row \rightarrow N \\
	pkValue(row(l, n, ps)) = n \\ \\
	 l \in Label, ps = Pair*, n \in N
\end{align*}
The value stored in concrete column of the row is represented by the pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		\inference{r = \varnothing}{\varnothing} \\
 		\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \in Rows, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}

>>>>

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented using an object-oriented language, which uses relational database as a data storage (as illustrated in the Figure \ref{fig:appStructure}) is object of interest of this paper. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software as a triple:
\begin{align*}
	\mathbf{software} = < Application, Database, \rho > \\
	sw : Applications \times Databases \times ORM \rightarrow Software \\ \\
	\mathbf{application} : Software \rightarrow Application \\
	application(sw(a, d, \rho)) = a \\ \\
	\mathbf{database} : Software \rightarrow Database \\
	database(sw(a, d, \rho)) = d \\
	a \lhd Application, d \lhd Database
\end{align*}
The software definition above introduce the first type used in this paper. The notation $a \lhd Application$ indicates the variable $a$ has the type $Application$.

The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user. The components of a software (application and database) are transformed likewise, thus the state is defined for them too.
\begin{align*}
 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s,  \\	s \lhd Software \vee s \lh Application \vee s \lhd Database
\end{align*}
Before the definition of the software consistency is provided, the notation of function transcription has to be introduced. Usually the definition of a function consists of two parts - an effect of the function and conditions when the effect occurs We decide to use following transcription, to simplify the notation and to improve understanding of functions:
\begin{align*}
f(x) =   \inference{conditions}{effect}
\end{align*}
The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
 	\mathbf{consistent} : State \rightarrow Boolean \\
 	consistent(s) = \begin{cases}
 		\inference{s = Consistent(s)}{true}\\\\
 		\inference{s = Inconsistent(s)}{false}\\
 \end{cases}\\
 s \lhd State
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
	\mathbf{state} : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ \\
\end{align*}

\subsection{Software Evolution}
\begin{figure}
\centering
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
	\label{fig:evolution}
\end{figure}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software and the function which changed the state of a software are called transformations. The evolution consists of change in application and in database as shown in the Figure \ref{fig:evolution}. We focus on situations when the evolution is initiated on the application level (as code refactoring), thus the transformation are designed to conform to the refactoring cases. However a case when the evolution begins on database level is also possible (eg. as a result of database optimization) it is not considered in this paper.

The evolution of the whole software has to be processed on the both software components - application and database. It means there is a set of application transformations called $AppT$ and a set of database transformations called $DbT$. Because there are needed different parameters for application and database evolution the evolution of database can not be derived from application evolution only. Thus the set of transformation called $E$ is defined to evolve the whole software together with mappings $\Psi$ between $E$ and $AppT$ and $\Phi$ between $E$ and $DbT$ is defined:
\begin{align*}
\Psi : E \rightarrow AppT \\
\Phi : E \rightarrow DbT
\end{align*}
The set $E$ containing the evolution transformations of a software mets our first goal - a single source of data evolution for the whole software.

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
$$\oplus : State \times E \rightarrow State $$
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
\circ : State \times AppT \rightarrow State \\
 \bullet : State \times DbT \rightarrow State
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.
\begin{align*}
 state(s) \oplus t = \begin{cases}
 	\inference{! consistent(state(s))}{u} \\ \\
 	\inference{consistent(state(s))}{state(sw(\Psi(s, t), \Phi(s,t))} 
\end{cases}\\ 
s \lhd Software, t \in E
\end{align*}
\begin{align*}
 state(a) \circ e = \begin{cases}
 	\inference{!\;consistent(state(a))}{a} \\ \\
 	\inference{consistent(state(a))}{e(a)} 
\end{cases}\\ 
a \lhd Application, e \in AppT
\end{align*}
\begin{align*}
 state(d) \bullet e = \begin{cases}
 	\inference{!\;consistent(state(d))}{d} \\ \\
 	\inference{consistent(state(d))}{e(d)} 
\end{cases}\\ 
d \lhd Application, e \in DbT
\end{align*}

\section{Software Model}
The model of an ORM software consists of three parts - an application model, a database model and an ORM. The model of application and database creates a type system for both ORM and evolutionary transformations.

The definitions of all parts assume a set $Label$ exists which contains all possible identifiers and each modeled entity is unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model.

\subsection{Application Model}
\label{sec:appModel}
The application model represents an application persistent layer created using object oriented language and functions which can alter the model. 
\subsubsection{Application Structure}
Each instance in an application is defined according to some class - the class defines the structure of its instances and relationships between them an the rest of the application. The instances are stored in the relational database in our case, therefore we define only the structure of classes in our model. 
\begin{align*}
  \mathbf{Application} = < Class* > \\ 
  \mathbf{app} : Class* \rightarrow  Application \\ \\
  \mathbf{classes} : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  \mathbf{class} : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label,a \lhd Applications, cs = Class* 
\end{align*}

\paragraph{Class} Class represents a basic structural unit in the application model. It has a unique name, one or more properties and it can be associated to other classes in the application. 
\begin{align*}
	\mathbf{Class} = <label, Property*, Association*> \\
	\mathbf{cl} : Label \times Property* \times Association* \rightarrow Class \\\\
	\mathbf{name} : Class \rightarrow Label \\
	name(cl(l, props, assocs)) = l \\ \\
	\mathbf{properties} : Class \rightarrow Property* \\
	properties(cl(l, props, assocs)) = props \\ \\
	\mathbf{associations} : Class \rightarrow Association* \\	associations(cl(l, props, assocs)) = assocs \\ 
	l \in Label, props \lhd Property*, assocs \lhd Association*
\end{align*}
Three more function are defined to obtain information about a class - function $primitives$ returns all properties of the class which represent a single value, whereas function $collections$ returns all properties representing collection of values. The last function called $associated$ determine if a class is associated by any other class in an application.
\begin{align*}
	\mathbf{primitives} : Class \rightarrow Property* \\	primitives(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
	\mathbf{collections} : Class \rightarrow Property* \\	collections(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
	\mathbf{associated} :  Label \times Application \rightarrow Boolean \\
  	associated(l,a) = \begin{cases}
  		\inference{\exists c \in classes(a) \wedge \exists e \in associations(c) : \\ reference(e) = l
  	}{ true } \\\\
 		false
 	\end{cases} \\ \\
	l \in Label, props = Property*, assocs = Association*,a \lhd Applications
\end{align*}

\paragraph{Property} Property represents a feature of a class which is represented as a primitive type. The property can be mandatory, can have a default value and according to its cardinality it can represent a single value or a collection of values. The $owningClass$ function returns the class in the model which owns the property.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
	\mathbf{owningClass} : Property \times Application \rightarrow Class  \\ 	
	owningClass(p, a) = \inference{\exists c \in classes(a) \wedge p \in properties(c)}{c}  \\ \\
	l \in Label, t \lhd AppType, val \in Values, n \in \mathbb{N^{*}}, m \in Boolean 
\end{align*}

\paragraph {Association} Associtation represents a connection between two classes. It has a unique name and the reference is represented by the label of referenced class. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of both association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\ \\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
	\mathbf{owningClass} : Association \times Application \rightarrow Class  \\
	owningClass(as, a) = \inference{\exists c \in classes(a) \wedge as \in associations(c)}{c}  \\ \\	
	l, ref \in Label,  n_1, n_2 \in \mathbb{N}, as \lhd Association
\end{align*}

@@

\paragraph{Application type} Application type ($AppType$) represents primitive type in the application. There are usually defined types such as String, Integer, Boolean etc. in contrast there is only one type in our model, because we focus on structural and data changes and type casting operations are not important for us. The only $AppType$ type is called $APPSTRING$.
\begin{align*}
\mathbf{AppType} = APPSTRING
\end{align*}

\subsubsection{Application consistency}

\subsection{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents stored instances. The database is defined as a tuple:
\begin{align*}
	\mathbf{Database} = < Table*, Row*> \\
	\mathbf{db} : Tables \times Rows \rightarrow Database \\ \\
	\mathbf{tables} : Database \rightarrow Tables \\
	tables(db(tabs, rs)) = tabs \\ \\
	\mathbf{rows} : Database \rightarrow Rows \\
	rows(db(tabs, rs)) = rs \\ \\
	tabs = Table*, rs = Row*, l \in Label,d \lhd Database
\end{align*}
\paragraph{Table} Table represents a basic concept of database schema. It has a unique name, one or more columns and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{align*}
	\mathbf{Table} = <label, primaryKey, Column*, ForeignKey*>\\
	\mathbf{tab} 	: Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
	\mathbf{name} : Table \rightarrow Label  \\
	name(tab(l, pk, cols, fks)) = l \\ \\
	\mathbf{primaryKey} : Table \rightarrow PrimaryKeys  \\
	primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
	\mathbf{columns} : Table \rightarrow Columns  \\
	columns(tab(l, pk, cols, fks)) = cols \\ \\
	\mathbf{foreignKeys} : Table \rightarrow ForeignKeys  \\
	foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk \lhd PrimaryKey, cols = Column*, fks = ForeignKey*
\end{align*}

\paragraph{Column} Column defines data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
	\mathbf{owningTable} : Column \times Database \rightarrow Table  \\
	owningTable(c, d) = \inference{\exists tab \in tables(d) \wedge c \in columns(tab)}{tab} \\ \\
l \in Label, t \lhd DbType, val \in Values,\\ cons \lhd Constraint*, c \lhd Column,d \lhd Database
\end{align*}

\paragraph{Foreign key} Foreign key is a reference to another table's primary key, it has an unique name and it can be constrained. The value of a foreign key is $\varnothing$ or a non-zero natural number.
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ 
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
	l, tRef \in Label, cons \lhd Constraint*,d \lhd Database, fk \lhd ForeignKey
\end{align*}

\paragraph{Primary key} Primary key is unambiguous identifier of a record in a table. The primary key is always provided (automatically generated) by the database as a non-zero natural number. Primary key is always defined with constraints $NOTNULL$ and $UNIQUE$. 
\begin{align*}
	\mathbf{PrimaryKey} =  < label > 	\\
	\mathbf{pk} : Label \rightarrow PrimaryKey
\end{align*}

\paragraph{Data Types} Database data types $DbType$ represents primitive types in the database. There are usually defined types such as Varchar, Integer, Boolean etc. There is only one type of column values defined in the model called $DBSTRING$. Next the $DBN$ type is defined for the values of keys.
\begin{align*}
	\mathbf{DbType} = DBSTRING | DBN
\end{align*}

\paragraph{Constraints} There are two types of constraints defined in the model. Both constraints are column constraints - first constraint defines non-empty columns, second constraint defines there has to be unique records in a column or foreign key.
\begin{align*}
	\mathbf{Constraint} = NOTNULL \; | \; UNIQUE 
\end{align*}

@@

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pkValue, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, n, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, n, ps)) = ps \\ \\
	\mathbf{pkValue} : Row \rightarrow N \\
	pkValue(row(l, n, ps)) = n \\ \\
	 l \in Label, ps = Pair*, n \in N
\end{align*}
The value stored in concrete column of the row is represented by a pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Value \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		\inference{r = \varnothing}{\varnothing} \\
 		\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \lhd Row*, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}
